Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-07 manuscript/manuscript.tex
ms updates

In our previous work, \textcite{Buffalo2019-io}, we did not investigate whether
background selection can lead to temporal autocovariance. Here, using
Forward-in-time simulations, we find that background selection can indeed
generate temporal autocovariance and lead to convergence correlations when
deleterious haplotypes are shared between populations and both removed by
selection.

\gc{COMMENT ON LARGER U than usual}

We simulated background selection in a 50 megabase region, where deleterious
alleles are randomly introduced by mutation. Following background selection
literature
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}, we
parameterize the mutation rate as the total number of deleterious mutations
introduced per diploid genome, per generation, and simulate values $U = \{0.5,
1.0, 1.5\}$. We also vary the strength of selection against the deleterious
mutations, $s = \{0.0, 0.01, 0.05, 0.1\}$ (where $s=0$ is a neutral control),
as well as different recombination rates ($r_{bp} = \{10^{-7}, 10^{-8}\}$).
Like other simulations, we burnin the population for $10N$ generations under
backgrounds selection. Overall, we find background selection does create
temporal covariance (Supplementary Material Figure
\ref{suppfig:supp-bgs-covs}), which are stronger under (1) higher deleterious
mutation rates and (2) larger selection coefficients. This latter point
initially seems at odds with background selection theory, as the level of
pairwise diversity in a region under strong background selection is invariant
with respect to the selection coefficient. However, looking at the background
selection $G(t)$ trajectories, we find that over time, the impact of linked
selection created by background selection appears to trend towards an
equilibrium in the $r_{BP} = 10^{-7}$ subfigure, and reaches an equilibrium in
the $r_{BP} = 10^{-8}$ subfigure that seems reasonably invariant to the choice
of $s$ (Supplementary Material Figure \ref{suppfig:supp-bgs-g-fix}). We believe
that these observations can be reconciled by $Cov(\Delta p_t,~ \Delta
p_{t^{\prime}})$ being larger for larger $s$ when $|t^{\prime}-t|$ is small,
but also decaying more rapidly with $|t^{\prime}-t|$, such that the overall
contribution of selection to allele frequency variance is invariant to $s$ (for
strong background selection). However, further future work is needed to fully
explore and understand the temporal covariance dynamics of background
selection.

@@

  \caption{The trajectories of $G(t)$ through time under background selection,
    under different recombination rates ($r_{BP}$, rows), selection
    coefficients ($s$), and deleterious mutation rates ($U$). The first column
    sets $U = 1.0$, and $s$ varies, while in the second column $s=0.05$ is held
  constant, and $U$ varies. Calculated using both
  including fixed sites (solid lines) and not including fixed sites (dashed
lines). \gc{MOVED THIS FIG UP} }

@@

%\begin{figure}[!ht]
 % \centering
  %\includegraphics[width=\textwidth]{figures/fig-bgs-G-without-fixations.pdf}

  %\caption{The trajectories of $G(t)$ through time under background selection,
   % under different recombination rates ($r_{BP}$, rows), selection
  %  coefficients ($s$), and deleterious mutation rates ($U$). The first column
  %  sets $U = 1.0$, and $s$ varies, while in the second column $s=0.05$ is held
 %% constant, and $U$ varies.}

 % \label{suppfig:supp-bgs-g}
%\end{figure}

@@

\gc{We found that the the results for background selection varied
  according to whether sites that fix or are lost through
time are included or excluded in the temporal covariance calculation can lead
to differences in the estimated $G(t)$temporal covariance. We discuss
this extensively in a subsequent section, Supplementary Materials Section \ref{supp:fixation}.}

%While investigating} whether different selection coefficients converged to
%the same levels of $G(t)$, we extended the number of generations we calculated
%our statistics on (e.g. in earlier exponential and Gaussian stabilizing
%selection fitness function, these were calculated to fifty generations; here
%we've calculated them to one hundred generations). This lead us to discover a
%subtle bias temporal covariance: whether sites that fix or are lost through
%ime are included or excluded in the temporal covariance calculation can lead
%to biases in temporal covariance. We discuss this extensively in the next
%section, Supplementary Materials Section \ref{supp:fixation}, but here we note
%that all the background selection results exclude fixed and lost sites; these
%are marked as missing and excluded from the temporal covariance and $G(t)$
%calculations.

@@

\subsubsection{Truncation Selection.}
We \vb{also} explored how directional truncation selection generates temporal
covariances. In these simulations we select the top \vb{10\%, 25\%, or 50\%} of
the phenotypic distribution of individuals to form the next generation. We
burnin these simulations \vb{using a neutral burnin routine, where trait
alleles are selectively neutral until directional selection is imposed}. More
extreme directional truncation selection generated larger initial covariances
(Figure \ref{suppfig:supp-trunc}B).  However, weaker truncation selection
generated more sustained positive covariances and so had a larger long-term
impact on the variance in allele frequency change. 
\gc{We noticed again strong effect fixation  or loss of sites
has on temporal covariance and $G(t)$ under truncation selection.}  Here,
since only a (potentially small) fraction of individuals contribute to the next
generation, sites can fix over very short timescales. Furthermore, the small
number of effective breeders contributing to the next generation shrinks $N_e$
considerably, which increases $\var(p_t - p_0$, the denominator of $G(t)$. We
see the effect of handling fixed/lost sites differently, and the faster rate of
drift in Supplementary Materials Figure \ref{suppfig:supp-trunc} (A), where
weaker truncation selection actually has higher levels of $G(t)$. Looking just
at temporal covariances, we find that stronger truncation selection (e.g.  a
smaller tail of individuals selected) does lead to greater temporal
covariances.

@@

\subsubsection{The Effect of Fixations in the empirical datasets}
\label{supp:fixation}
\gc{In our simulation results, we noticed the temporal covariances and $G(t)$ statistics
can differ depending on how allele frequencies of zero or one are handled.

%In analyzing both the empirical datasets included in our study and our
%simulation results, we noticed the temporal covariances and $G(t)$ statistics
%can differ depending on how allele frequencies of zero or one are handled.
Generally, temporal covariances should be calculated on polymorphic sites; once
a site has reached fixation or loss, its allele frequency change $\Delta p_t =
0$ and including these sites in the temporal covariance calculation can lead to
biases. However, with \emph{sample} allele frequencies,
rather than population frequencies, a site with observed frequency zero or one
may still be segregating, but by chance not sampled at a timepoint. Here, we
discuss the effect of including sites with frequency zero or one, and show our
empirical results are not qualitatively different when analyzed excluding fixed
sites.

With empirical data calculated on sample allele
 frequencies, low frequency minor alleles many not be sampled at some
 timepoints, and excluding these observations (instead of treating it as a
 trajectory that has a 0 frequency timepoint) biases estimates of
 quantities such as $N_e$ as it biased the analysis towards
 intermediate frequency alleles.
 Additionally, we tried only dropping fixed or loss sites from the temporal
 covariance calculations that were at the end or the beginning of a trajectory
 (e.g., as if the site was created by a new mutation or fixed); while this
 ameliorated some of this bias it did not remove all of it. 
 Overall, we found by trying all these approaches that not removing fixed or lost sites was the best way to deal with
 sample allele frequencies that could be missing from some timepoints. }

% We noticed in some simulations, when temporal covariances are calculated over
% longer time periods (i.e. $>50$ generations), that neutral simulations appear
% to slowly trend towards negative $G(t)$ (Supplementary Materials Figure
% \ref{suppfig:supp-bgs-g-fix}). In investigating this bias, we found that the
% effect went away when we marked fixed or lost sites as missing, and calculated
% the temporal covariance; furthermore, this affected the $G(t)$ trajectories
% under selection too (Supplementary Material Figure
% \ref{suppfig:supp-bgs-g-fix}).

% Looking a neutral simulations results, we found that spuriously significant
% $G(t)$ trajectories (e.g. different from zero under the case of neutrality) can
% be created by conditioning on sojourn time. Our intuition, which will need to
% be confirmed by future work and theory, is that by conditioning on a quick
% sojourn time to fixation (or loss) from an initial intermediate frequency,
% leads to trajectories that, purely by chance, have many upward (or downward)
% allele frequency changes, which leads to a positive sum of pairwise temporal
% covariances (the numerator of $G(t)$). Similarly, conditioning on long sojourn
% times implies that frequency changes alternate back and forth quite a bit, such
% the sum of pairwise temporal covariances could be slightly negative. However,
% the overall affect could also be affected by the average allele frequency
% change across loci and other factors.

% To further complicate the matter, our simulation allele frequencies are
% population, rather than sample frequencies, when an allele is fixed or lost is
% known with certainty. With empirical data calculated on sample allele
% frequencies, low frequency minor alleles many not be sampled at some
% timepoints, and excluding these observations (instead of treating it as a
% trajectory that has a 0 frequency timepoint) biases estimates.  We observed
% this bias created by ignoring fixed or loss sites by comparing our $N_e$
% estimates for the \textcite{Barghi2019-qy} study (estimated from the total
% variance in allele frequency change) with the estimates from the original
% paper. We found that ignoring fixed or lost sites (by treating them as missing
% data in calculating the covariances) led to many low-frequency alleles not
% contributing to the variance, leading this statistic to be calculated on more
% intermediate frequency alleles and thus biasing the $N_e$ estimate downwards.
% Additionally, we tried only dropping fixed or loss sites from the temporal
% covariance calculations that were at the end or the beginning of a trajectory
% (e.g., as if the site was created by a new mutation or fixed); while this
% ameliorated some of this bias, it still did not lead to $N_e$
% estimates  GUESS I"M UNSURE WHY THIS IS A CRITERIA FOR BEING CORRECT
% congruent with the original studies. Overall, we found by trying all these
% approaches that not removing fixed or lost sites was the best way to deal with
% sample allele frequencies that could be missing from some timepoints.

@@

To ensure that our findings were robust to handling sites with a frequency of
zero or one differently, we regenerated Figures \ref{fig:figure-1} and Figure
\ref{fig:figure-3} but excluded frequencies of zero or one. Specifically, we
wanted to ensure that our finding that later timepoints had negative
covariances at later timepoints was not spuriously caused by the way fixed or
lost sites are handled. We see no qualitative difference (Supplementary
Material Figures \ref{suppfig:supp-fig-1-nofix} and
\ref{suppfig:supp-fig-3-nofix}) that emerges when sites with frequency zero or
one are excluded.

% \subsubsection{The Effect of Fixations}
% \label{supp:fixation}

% In analyzing both the empirical datasets included in our study and our
% simulation results, we noticed the temporal covariances and $G(t)$ statistics
% can differ depending on how allele frequencies of zero or one are handled.
% Generally, temporal covariances should be calculated on polymorphic sites; once
% a site has reached fixation or loss, its allele frequency change $\Delta p_t =
% 0$ and including these sites in the temporal covariance calculation can lead to
% biases, which we discuss below. However, with \emph{sample} allele frequencies,
% rather than population frequencies, a site with observed frequency zero or one
% may still be segregating, but by chance not sampled at a timepoint. Here, we
% discuss the effect of including sites with frequency zero or one, and show our
% empirical results are not qualitatively different when analyzed excluding fixed
% sites.

% \begin{figure}[!ht]
%   \centering
%   \includegraphics[width=\textwidth]{figures/fig-bgs-G-with-fixations.pdf}

%   \caption{$G(t)$ trajectories from background selection simulations both
%   including fixed sites (solid lines) and not including fixed sites (dashed
% lines). The dashed lines are identical to those in Supplementary Materials
% Figure \ref{suppfig:supp-bgs-g}. }

%   \label{suppfig:supp-bgs-g-fix}
% \end{figure}

% We noticed in some simulations, when temporal covariances are calculated over
% longer time periods (i.e. $>50$ generations), that neutral simulations appear
% to slowly trend towards negative $G(t)$ (Supplementary Materials Figure
% \ref{suppfig:supp-bgs-g-fix}). In investigating this bias, we found that the
% effect went away when we marked fixed or lost sites as missing, and calculated
% the temporal covariance; furthermore, this affected the $G(t)$ trajectories
% under selection too (Supplementary Material Figure
% \ref{suppfig:supp-bgs-g-fix}).

% Looking a neutral simulations results, we found that spuriously significant
% $G(t)$ trajectories (e.g. different from zero under the case of neutrality) can
% be created by conditioning on sojourn time. Our intuition, which will need to
% be confirmed by future work and theory, is that by conditioning on a quick
% sojourn time to fixation (or loss) from an initial intermediate frequency,
% leads to trajectories that, purely by chance, have many upward (or downward)
% allele frequency changes, which leads to a positive sum of pairwise temporal
% covariances (the numerator of $G(t)$). Similarly, conditioning on long sojourn
% times implies that frequency changes alternate back and forth quite a bit, such
% the sum of pairwise temporal covariances could be slightly negative. However,
% the overall affect could also be affected by the average allele frequency
% change across loci and other factors.

% To further complicate the matter, our simulation allele frequencies are
% population, rather than sample frequencies, when an allele is fixed or lost is
% known with certainty. With empirical data calculated on sample allele
% frequencies, low frequency minor alleles many not be sampled at some
% timepoints, and excluding these observations (instead of treating it as a
% trajectory that has a 0 frequency timepoint) biases estimates.  We observed
% this bias created by ignoring fixed or loss sites by comparing our $N_e$
% estimates for the \textcite{Barghi2019-qy} study (estimated from the total
% variance in allele frequency change) with the estimates from the original
% paper. We found that ignoring fixed or lost sites (by treating them as missing
% data in calculating the covariances) led to many low-frequency alleles not
% contributing to the variance, leading this statistic to be calculated on more
% intermediate frequency alleles and thus biasing the $N_e$ estimate downwards.
% Additionally, we tried only dropping fixed or loss sites from the temporal
% covariance calculations that were at the end or the beginning of a trajectory
% (e.g., as if the site was created by a new mutation or fixed); while this
% ameliorated some of this bias, it still did not lead to $N_e$ estimates
% congruent with the original studies. Overall, we found by trying all these
% approaches that not removing fixed or lost sites was the best way to deal with
% sample allele frequencies that could be missing from some timepoints.

% \begin{figure}[!ht]
%   \centering
%   \includegraphics[width=0.6\textwidth]{figures/figure-1-G-covs-nofix.pdf}

%   \caption{The effect of excluding fixed/lost sites in the calculation of the
%   temporal covariances and $G(t)$ trajectories of the \textcite{Barghi2019-qy}
% data. Dashed lines are those including fixed/lost sites (i.e. the original
% Figure 1), and solid lines are excluding fixed/lost sites.}

%   \label{suppfig:supp-fig-1-nofix}
% \end{figure}

% \begin{figure}[!ht]
%   \centering
%   \includegraphics{figures/figure-3-hists-b-without-fixations.pdf}

%   \caption{A version of Figure \ref{fig:figure-3} (A) and (B) excluding fixed
%     and lost sites. The same qualitative pattern holds as the original figure,
%     which did not exclude fixed and lost sites: there is an enrichment of
%     positive temporal covariances between near timepoints ($k=2$) in the
%   \textcite{Barghi2019-qy} study, and an excess of negative temporal
% covariances at more distant timepoints ($k=2$). }

%   \label{suppfig:supp-fig-3-nofix}
% \end{figure}

% To ensure that our findings were robust to handling sites with a frequency of
% zero or one differently, we regenerated Figures \ref{fig:figure-1} and Figure
% \ref{fig:figure-3} but excluded frequencies of zero or one. Specifically, we
% wanted to ensure that our finding that later timepoints had negative
% covariances at later timepoints was not spuriously caused by the way fixed or
% lost sites are handled. We see no qualitative difference (Supplementary
% Material Figures \ref{suppfig:supp-fig-1-nofix} and
% \ref{suppfig:supp-fig-3-nofix}) that emerges when sites with frequency zero or
% one are excluded.

% \begin{figure}[!ht]
%   \centering
%   \includegraphics[width=\textwidth]{figures/fig-both-trunc.pdf}

%   \caption{$G(t)$ trajectories (A) and temporal covariances (B) from
%     truncation selection simulations for different numbers of individuals
%     selected (line color). Dashed lines indicate $G(t)$ trajectories and
%     temporal covariances calculated \emph{including} fixed sites, while the
%     solid lines exclude fixed sites. All values are averaged over 30 replicate
%     simulations; the lines in the right figure are loess smoothed, while points
%     are averages. The solid lines of the temporal covariances have been
%     excluded in the left figure for clarity, but are similar except they do not
%     become negative.}

%     \label{suppfig:supp-trunc}
% \end{figure}

% Through simulations, we noticed the strongest effect fixation  or loss of sites
% has on temporal covariance and $G(t)$ is under truncation selection. Here,
% since only a (potentially small) fraction of individuals contribute to the next
% generation, sites can fix over very short timescales. Furthermore, the small
% number of effective breeders contributing to the next generation shrinks $N_e$
% considerably, which increases $\var(p_t - p_0$, the denominator of $G(t)$. We
% see the effect of handling fixed/lost sites differently, and the faster rate of
% drift in Supplementary Materials Figure \ref{suppfig:supp-trunc} (A), where
% weaker truncation selection actually has higher levels of $G(t)$. Looking just
% at temporal covariances, we find that stronger truncation selection (e.g.  a
% smaller tail of individuals selected) does lead to greater temporal
% covariances.

>>>>

In our previous work, \textcite{Buffalo2019-io}, we did not investigate whether
background selection can lead to temporal autocovariance. Here, using
forward-in-time simulations, we find that background selection can indeed
generate temporal autocovariance and lead to convergence correlations when
deleterious haplotypes are shared between populations and both removed by
selection.

We simulated background selection in a 50 megabase region, where deleterious
alleles are randomly introduced by mutation. Following background selection
literature
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}, we
parameterize the mutation rate as the total number of deleterious mutations
introduced per diploid genome, per generation, and simulate values $U = \{0.5,
  1.0, 1.5\}$. \vb{Note that for our 50Mb region, our choice of BGS $U$
  parameters are on the higher end of the spectrum expected for
  \emph{Drosophila}, but we wanted to ensure the strength was sufficient to see
  a signal. Note that if $U \approx 1.6$ \parencite{Elyashiv2016-vt}, and the
\emph{Drosophila} genome is $\approx 140$Mb, our region is $\approx 36\%$ of
the genome; this implies a reasonable $U$ for our region is $U \approx 0.57$,
which is close to the bottom end of our parameter range.} We also vary the
strength of selection against the deleterious mutations, $s = \{0.0, 0.01,
0.05, 0.1\}$ (where $s=0$ is a neutral control), as well as different
recombination rates ($r_{bp} = \{10^{-7}, 10^{-8}\}$).  Like other simulations,
we burnin the population for $10N$ generations under backgrounds selection.
Overall, we find background selection does create temporal covariance
(Supplementary Material Figure \ref{suppfig:supp-bgs-covs}), which are stronger
under (1) higher deleterious mutation rates and (2) larger selection
coefficients. This latter point initially seems at odds with background
selection theory, as the level of pairwise diversity in a region under strong
background selection is invariant with respect to the selection coefficient.
However, looking at the background selection $G(t)$ trajectories, we find that
over time, the impact of linked selection created by background selection
appears to trend towards an equilibrium in the $r_{BP} = 10^{-7}$ subfigure,
and reaches an equilibrium in the $r_{BP} = 10^{-8}$ subfigure that seems
reasonably invariant to the choice of $s$ (Supplementary Material Figure
\ref{suppfig:supp-bgs-g-fix}). We believe that these observations can be
reconciled by $Cov(\Delta p_t,~ \Delta p_{t^{\prime}})$ being larger for larger
$s$ when $|t^{\prime}-t|$ is small, but also decaying more rapidly with
$|t^{\prime}-t|$, such that the overall contribution of selection to allele
frequency variance is invariant to $s$ (for strong background selection).
However, further future work is needed to fully explore and understand the
temporal covariance dynamics of background selection.

@@

  \caption{The trajectories of $G(t)$ through time under background selection,
    under different recombination rates ($r_{BP}$, rows), selection
    coefficients ($s$), and deleterious mutation rates ($U$). The first column
    sets $U = 1.0$, and $s$ varies, while in the second column $s=0.05$ is held
  constant, and $U$ varies. Calculated using both including fixed sites (solid
lines) and not including fixed sites (dashed lines).}

@@

@@

\vb{Finally, we found in these background selection results that including or
excluding sites fixed or lost through time can lead  to differences in the
estimated $G(t)$ trajectories and temporal covariance. We discuss this extensively in a
subsequent section, Supplementary Materials Section \ref{supp:fixation}.}

@@

\subsubsection{Truncation Selection.}

We \vb{also} explored how directional truncation selection generates temporal
covariances. In these simulations we select the top \vb{10\%, 25\%, or 50\%} of
the phenotypic distribution of individuals to form the next generation. We
burnin these simulations \vb{using a neutral burnin routine, where trait
alleles are selectively neutral until directional selection is imposed}. More
extreme directional truncation selection generated larger initial covariances
(Figure \ref{suppfig:supp-trunc}B).  However, weaker truncation selection
generated more sustained positive covariances and so had a larger long-term
impact on the variance in allele frequency change.  We again noticed strong
effect fixation  or loss of sites has on temporal covariance and $G(t)$ under
truncation selection.  Here, since only a (potentially small) fraction of
individuals contribute to the next generation, sites can fix over very short
timescales. Furthermore, the small number of effective breeders contributing to
the next generation shrinks $N_e$ considerably, which increases $\var(p_t -
p_0)$, the denominator of $G(t)$. We see the effect of handling fixed/lost
sites differently, and the faster rate of drift in Supplementary Materials
Figure \ref{suppfig:supp-trunc}A, where weaker truncation selection actually
has higher levels of $G(t)$. Looking just at temporal covariances, we find that
stronger truncation selection (e.g.  a smaller tail of individuals selected)
does lead to greater temporal covariances. \vb{Overall, these truncation
  selection simulations demonstrate the value of considering both absolute
  measures of selection's effect on allele frequency changes, i.e. temporal
  covariance, as well as measures relative to drift, i.e. the $G(t)$
  trajectories. While selecting a smaller tail of individuals is associated
  with \emph{stronger} selection, it also is leading to a faster rate of drift,
which can distort conclusions inferred from considering the $G(t)$ trajectories
alone.}

@@

\subsubsection{The Effect of Fixations in the empirical datasets}
\label{supp:fixation}

In our simulation results, we noticed the temporal covariances and $G(t)$
statistics can differ depending on how allele frequencies of zero or one are
handled. Generally, temporal covariances should be calculated on polymorphic
sites; once a site has reached fixation or loss, its allele frequency change
$\Delta p_t = 0$ and including these sites in the temporal covariance
calculation can lead to biases. However, with \emph{sample} allele frequencies,
rather than population frequencies, a site with observed frequency zero or one
may still be segregating, but by chance not sampled at a timepoint. Here, we
discuss the effect of including sites with frequency zero or one, and show our
empirical results are not qualitatively different when analyzed excluding fixed
sites.

With empirical data calculated on sample allele frequencies, low frequency
minor alleles may not be sampled at some timepoints, and excluding these
observations (instead of treating it as a trajectory that has a 0 frequency
timepoint) biases estimates of quantities such as $N_e$ towards intermediate
frequency alleles. Additionally, we tried only dropping fixed or loss sites
from the temporal covariance calculations that were at the end or the beginning
of a trajectory (e.g., as if the site was created by a new mutation or fixed);
while this ameliorated some of this bias it did not remove all of it. Overall,
we found by trying all these approaches that not removing fixed or lost sites
was the best way to deal with sample allele frequencies that could be missing
from some timepoints.

@@

To ensure that our findings were robust to handling sites with a frequency of
zero or one differently, we regenerated Figures \ref{fig:figure-1} and Figure
\ref{fig:figure-3} but excluded frequencies of zero or one. Specifically, we
wanted to ensure that our finding that \vb{temporal covariances were negative
at later timepoints} was not spuriously caused by the way fixed or lost sites
are handled. We see no qualitative difference (Supplementary Material Figures
\ref{suppfig:supp-fig-1-nofix} and \ref{suppfig:supp-fig-3-nofix}) that emerges
when sites with frequency zero or one are excluded.
